---
title: "Statistical Inference with Linear Models: Housing Loan Applications in Fullerton"
author: "Your Name Here!"
format: html
editor: visual
---

## Load Packages, Import Data, Massage Data

```{r}
#| label: load packages
library(tidyverse)
library(infer)
library(rstanarm)
library(broom.mixed)
```

Import the loans_OC dataset as `loans`:

```{r}
#| label: import data 

```

We will work with the same relationships we looked at in the Summarizing Linear Models activity:

```{r}
#| label: summarize each tract

loans_fullerton2 <- loans |>
  filter(city == "Fullerton") |>
  mutate(
    loan_amount_1000 = loan_amount/1000,
    property_value_1000 = property_value/1000,
    loan_to_value_ratio = loan_amount_1000/property_value_1000)

fullerton_tracts <- loans_fullerton2 |>
  group_by(census_tract) |>
  summarize(
    applications = n(),
    population = mean(tract_population),
    minority = mean(tract_minority_population_percent),
    income = mean(tract_to_msa_income_percentage),
    median_value = median(property_value_1000, na.rm = TRUE)
  )
```

## Permutation Testing for Property Values vs. Minority

```{r}
#| label: scatterplot of median property values vs minority

ggplot(data = fullerton_tracts,
       mapping = aes(
         x = minority,
         y = median_value
       )) +
  geom_point() +
  geom_smooth(method = "lm", color = "darkblue", 
              linetype = "dashed", se = FALSE) +
  labs(
    title = "Housing Values in Fullerton Census Tracts",
    x = "Percentage of Minorities in Tract",
    y = "Median Property Value (1000s of $)"
  )

```

First let's do a permutation test for the significance of percent minority as a predictor of median value.

Our null model is

$$
median\_value = \beta_0 + ERROR
$$

implying that our null hypothesis is that percentage of minorities in the tract is unrelated to median property value. Our proposed model is

$$
median\_value = \beta_0 + \beta_1 \times minority + ERROR
$$

We'll perform all of the steps of the hypothesis testing procedure chained together using the pipe `|>`:

```{r}
#| label: permutation test medvalue minority

set.seed(284)
lm_medvalue_minority_ht <- fullerton_tracts |>
  specify(median_value ~ minority) |>
  hypothesize(null = "independence") |>
  generate(type = "permute", reps = 1000) |> 
  calculate(stat = "slope")

```

Here we're going to use slightly different syntax to find the observed value of the slope.

```{r}
#| label: calculate observed slope with infer

obs_slope <- fullerton_tracts |>
  specify(median_value ~ minority) |>
  calculate(stat = "slope")

obs_slope
```

Now we can plot our null distribution. Notice the slight difference in how we created the red vertical line.

```{r}
#| label: null distribution medvalue minority

ggplot(
  data = lm_medvalue_minority_ht,
  mapping = aes(x = stat)
) +
  geom_histogram(
  ) +
  geom_vline(
    xintercept = obs_slope$stat,
    color = "red",
    linewidth = 1.25
  ) +
  labs(
    x = "Slope of Regression Line",
    y = "Number of Fake Samples"
  )
```

To get the p-value, we can just use the function `get_p_value` from the `infer` package. This is the whole reason why we got the observed slope out a bit differently.

```{r}
#| label: get p-value medvalue minority

get_p_value(
  lm_medvalue_minority_ht,
  obs_stat = obs_slope,
  direction = "two.sided" # I prefer the ., but _ or a space works just as well
)
```

### Your Turn

1.  Using a 5% significance level, does percent minority appear to have a (statistically) significant effect on median property value? Why or why not?

## Permutation Testing for Property Values vs. Income

### Your Turn

1.  Perform a permutation test to determine whether a model predicting median property value from `income` is a significantly better fit than the null model. Include all steps:

    a.    Write out the proposed model and the null model
    
    b.    Perform the resampling procedure to generate a null distribution of the slope of the regression line
  
    c.    Plot your null distribution and indicate the observed value of the slope on the graph
    
    d.    Obtain the p-value
    
    e.    Determine whether `income` has a significant effect on median property value

## Bayesian Inference for Property Values vs. Minority

In the Bayesian framework, the slope $\beta_1$ of the population model is treated as a variable with some uncertainty. We will use our regular weakly informative prior that the true slope is "somewhere around 0".

```{r}
#| label: fit Bayesian SLR

bayeslm_medvalue_minority <- stan_glm(
  median_value ~ minority,
  data = fullerton_tracts,
  prior = normal(location = 0),
  seed = 1 # seed for reproducibility
)

plot(
  bayeslm_medvalue_minority,
  pars = "minority",
  plotfun = "hist",
  binwidth = 1
) +
  labs(
    title = "Posterior distribution",
    x = "Slope corresponding to percent minority",
    y = "Number of simulations"
  ) +
  theme(
    plot.title = element_text(size = 28),
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )

```

Let's use this posterior distribution to do some Bayesian inference:

```{r}
#| label: Bayesian Inference medvalue minority

bayeslm_medvalue_minority |> 
  posterior_interval(
    pars = "minority",
    prob = 0.8
    )
```

### Your Turn

1. Write a sentence to interpret this 80% Bayesian uncertainty interval.

2. Explain why the posterior distribution looks different from the null distribution. What specifically does the x-axis represent on the posterior distribution, and how is that different from what the x-axis represents on the null distribution?

## Bayesian Inference for Property Values vs. Income

### Your Turn

1. Obtain an 80% Bayesian uncertainty interval for the slope of the population model relating `median_value` and `income`.
