---
title: "Model Selection for Classification: Housing Loan Applications in Orange County"
format: html
editor: visual
---

## Load Packages, Import Data, Massage Data

```{r}
#| label: do this first

here::i_am("Activities/Model-Selection-Classification-OC-Mortgages.qmd")

library(tidyverse)
library(rsample)
library(janitor)
library(broom)
library(yardstick)
library(probably)

loans <- read_csv(here::here("Data/loans_OC.csv"))
```

To massage the data, we'll start by getting only the loans for housing in Fullerton with a known `property_value`. We'll then create `loan_amount_1000` and `property_value_1000` (as we've done in previous activities) and convert `loan_to_value_ratio` into percent so we don't make a one-unit increase larger than the range of the variable. We'll also explicitly convert `action` to a factor variable because we need to do that for our logistic regression models, and we'll define the reference level for `ethnicity`.

```{r}
#| label: get only loans from fullerton

loans_fullerton <- loans |>
  filter(city == "Fullerton",
         !is.na(property_value)) |>
  mutate(
    loan_amount_1000 = loan_amount/1000,
    property_value_1000 = property_value/1000,
    loan_to_value_ratio = loan_amount_1000/property_value_1000*100,
    action = action |>
      as.factor(), # convert to factor
    ethnicity = ethnicity |>
      as.factor() |> # convert ethnicity to factor
      relevel(ref = "Not Hispanic or Latino")
)
```

Now we'll do our training/test split.

```{r}
#| label: initial split
set.seed(15858)
loans_split <- loans_fullerton |>
  initial_split(
    strata = action,
    prop = 0.80
  )

loans_train <- training(loans_split)
loans_test <- testing(loans_split)
```


## Setting Up our Cross-Validation

We create our folds using the training set. We'll use 5-fold cross-validation.

```{r}
#| label: v-fold cv setup

```

## Feature Selection

1. Define four different subsets of predictors to investigate in your logistic regression model. Consider including:

-   loan amount, property value, and/or loan-to-value ratio
-   demographic information (age, sex, ethnicity, race): note that if you choose race we might have to think about how to recode it
-   the census tract, or information about it

## Cross-Validation

### Create the Function

We first create our function that's going to fit all models on all but one fold and make predictions on the other fold.

```{r}
#| label: function for prediction

loans_prediction <- function(split){
  # our input is the cross-validated split
  
  train <- analysis(split)
  valid <- assessment(split)
  
  ## glm null model
  glm0 <- glm(action ~ 1, data = train, family = "binomial")
  ## Add more GLMs here
  
  valid_predictions <- valid |>
    mutate(
      prob0 = predict(glm0, newdata = valid, type = "response"),
    ## Add more probabilities here

    )
  
  return(valid_predictions)
}
```

### Map the Function Over All Folds

```{r}
#| label: map the predictions

```

### Investigate Model Performance

First we need to figure out whether we are predicting the probability of Approved or Not approved. The easiest way to do this is with the contrasts function:

```{r}
#| label: what are we predicting


```

Since "Not approved" is a 1, we are predicting the probability of not approved.

However, to obtain the Brier scores, we need to give the `brier_class` function the probability of being Approved (since that's the first level).

```{r}
#| label: all predictions

```

1. Why is the mean Brier score so low for even the null model?

### Investigate Model Calibration

Investigating model calibration is a bit more difficult with classification models because we need to think about how the "observed proportion" works. We can use some plotting functions in the `probably` package to think about this automatically:

```{r}
#| label: event rate vs prediction window plot

```

Note that there is some weirdness here that I haven't been able to fully figure out. Part of the issue is that the default ranges are 0-0.10, 0.10-0.20, etc. and there isn't a clear way to customize these ranges. 

## Select a Model

1. Select one of your models as the final model, fit it on the entire training set, and then predict the probability of being approved/not approved for each application in the test set.

```{r}
#| label: final model


```

## Evaluate the Final Model

We don't know what decision threshold our actual decision maker will be using, so one reasonable idea is to investigate the model performance across the entire range of possible decision thresholds.

1.  Create and plot a receiver operating characteristic (ROC) curve for the test set predictions.

```{r}
#| label: create ROC Curve as tibble

```

```{r}
#| label: plot ROC curve

```

2. How do we read this plot? What does `sensitivity` indicate about the predictions? What does `1 - specificity` indicate about the predictions?
