---
title: "Fitting and Summarizing Linear Models: Rent Prices in Southern California (Part 2)"
author: "Your Name Here!"
format: html
editor: visual
---

In this activity, our goal in this activity is to build a model to predict the listed rent price of a house (in 2022).

The SoCalRent1 dataset contains information about 183 houses or condos available to rent within a "reasonable distance" of Fullerton in May 2022.

The SoCalRent2 dataset contains information about an additional 117 houses or condos available to rent during the same time period.

See the SoCalRent data dictionary for descriptions of the variables.

## Load Packages, Import Data, Massage Data

```{r}
#| label: do this first
here::i_am("Activities/Multiple-Linear-Regression-SoCalRent.qmd")

library(tidyverse)
library(broom)
library(yardstick)
library(rstanarm)

SoCalRent1 <- read_csv(here::here("Data/SoCalRent1.csv"))
SoCalRent2 <- read_csv(here::here("Data/SoCalRent2.csv"))
```

```{r}
#| label: filter SoCalRent1 to get SoCalRent_train

```

## Building the Model

We will now build a *multiple* regression model, indicating we are predicting price using more than one predictor variable. Here we will use `SqFt`, `Beds`, and `Baths` as predictor variables.

```{r}
#| label: build model including Beds and Baths

```

### Investigating the Model

1.  Use the output to write out the equation of the least-squares regression line.

2.  Write a sentence to interpret the slope corresponding to `Beds` in the model.

3.  Does this interpretation make sense in the context of the collected data? Why or why not?

### Making Predictions

1.  Perform the same `filter`ing on the `SoCalRent2` dataset that you did on the `SoCalRent1` dataset. Call this new dataset `SoCalRent_test`.

2.  Using the `augment` function, predict the rent price of every house in the `SoCalRent_test` dataset.

3. Obtain the mean absolute error (MAE) and root-mean-squared error (RMSE) for the `SoCalRent_test` dataset. Do these numbers suggest anything concerning about the accuracy of the predictions? Why or why not?

4.  Produce a histogram of the prediction errors (residuals) for the `SoCalRent_test` dataset. Briefly describe the distribution of the residuals. Does this histogram suggest anything concerning about the accuracy of the predictions? Why or why not?

5.  Produce a scatterplot with the predicted (fitted) prices on the x-axis and the residuals on the y-axis. Add a linear and/or nonlinear trend to the plot (using `geom_smooth`). Does this scatterplot suggest anything concerning about the accuracy of the predictions? Why or why not?


## Bayesian Modeling

When we make a Bayesian model, we need to put a prior distribution on each coefficient in the model. We will use the `autoscale = TRUE` argument to tell R to pick better priors for the `scale` (standard deviation of the prior distribution).

1. What would be a reasonable "prior guess" for the slope relating Price and each of the three predictors in the model? We will use these guesses as the `location` (mean) of our `normal` prior distribution.

```{r}
#| label: bayesian model

```

2. Create a set of three histograms showing the posterior distributions of the slopes corresponding to `SqFt`, `Beds`, and `Baths`.

```{r}
#| label: histograms of posterior distributions

```

3. Obtain 90% uncertainty intervals for the slopes.

```{r}
#| label: uncertainty intervals

```

### Making Predictions the Bayesian Way

When we are predicting rent price, we (in theory) know the square footage, number of bedrooms, and number of baths, but we are uncertain about the price.

We can use *each one* of the 4000 simulated "possible population models" to make a prediction for the test set.

```{r}
#| label: create SoCalRent_test from SoCalRent2


```

```{r}
#| label: make predictions with bayeslm

```

Notice: each column represents one of the houses in the `SoCalRent_test` dataset and each row represents one of the 4000 population models.

```{r}
#| label: plot posterior predictive distribution for first house


```

1. Does anything look weird about these predictions? 

2. Obtain posterior predictive intervals for each house in the `SoCalRent_test` dataset. Determine whether each interval contains the true Price. 

```{r}
#| label: posterior predictive intervals


```

3. For which houses does our model "miss" with the posterior predictive intervals? Do they have anything in common that might give insight into a better model?

4. How does the proportion of "correctly predicted" rent prices compare to probability chosen for the posterior predictive intervals?

5. Produce a plot comparing the posterior predictive intervals to the actual rent prices in the test set.

```{r}
#| label: plot ppis with geom_segment

```
