---
title: "Statistical Inference: Rent Prices in Southern California"
format: html
editor: visual
---

In this activity, we will do (frequentist) statistical inference for the SoCalRent data.

## Load Packages, Import Data, Massage Data

```{r}
#| label: do this first
here::i_am("Activities/Summarizing-Linear-Models-SoCalRent.qmd")

library(tidyverse)
library(infer)

SoCalRent1 <- read_csv(here::here("Data/SoCalRent1.csv"))
```

In the "Summarizing Linear Models" activity, we removed some of the houses from this dataset. Repeat that now:

```{r}
#| label: create SoCalRent_train


```

## Permutation Testing

Our goal in this part of the activity is to determine whether a linear model predicting the listed rent price of a house (in 2022) based on its square footage explains "significantly more" of the variability in rent price than a null model.

### Step 1: Define the Models

Our null model is

$$
\text{Price} = \beta_0 + \text{ERROR}
$$

Our proposed population model is

$$
\text{Price} = \beta_0 + \beta_1 \times \text{SqFt} + ERROR
$$

### Step 2: Observed Value of the Test Statistic

Let's find the observed R-squared value using techniques in the `infer` package:

```{r}
#| label: observed R-squared SqFt

```

### Step 3: Simulated Values of the Test Statistic

Now let's generate the null distribution:

```{r}
#| label: permutation test  - Price vs SqFt

set.seed(284)

```

### Step 4: Compare the Observed and Simulated Values

Now we graph the null distribution:

```{r}
#| label: simulated null distribution - Price vs SqFt

```

And we compute the p-value. In this case it's quite obvious what the p-value should be:

```{r}
#| label: compute p-value


```

Remember that this isn't actually a p-value of 0 (because this observed sample is *technically* one of the samples we could have observed by doing our random swapping), but it's so small that our model explains "significantly more" of the variability in price than the null model does.

## Confidence Intervals

Our goal in this part of the activity is to determine the average increase in rent price for every additional square foot.

### Step 1: Define the Model

Our proposed population model is

$$
\text{Price} = \beta_0 + \beta_1 \times \text{SqFt} + ERROR
$$

Thus, our goal is to estimate $\beta_1$.

### Step 2: Obtain Simulated Samples

```{r}
#| label: bootstrap distribution - Price vs SqFt

set.seed(284)


```

### Step 3: Calculate the Confidence Interval

```{r}
#| label: bootstrap distribution of slope


```

```{r}
#| label: confidence interval

```

```{r}
#| label: infer visualize

```

### Step 4: Interpret the Confidence Interval

With 95% confidence, we estimate that the population mean rent price increases by between $\$1.07$ and $\$2.29$ per square foot. 