---
title: "Fitting and Summarizing Linear Models: Housing Loan Applications in Fullerton"
author: "Your Name Here!"
format: html
editor: visual
---

## Load Packages, Import Data, Massage Data

```{r}
#| label: load packages
library(tidyverse)
library(broom)
```

Import the loans_OC dataset as `loans`:

```{r}
#| label: import data 

```

We will work with the scatterplots we created at the end of the Summarizing Numerical Data activity:

```{r}
#| label: summarize each tract

loans_fullerton2 <- loans |>
  filter(city == "Fullerton") |>
  mutate(
    loan_amount_1000 = loan_amount/1000,
    property_value_1000 = property_value/1000,
    loan_to_value_ratio = loan_amount_1000/property_value_1000)

fullerton_tracts <- loans_fullerton2 |>
  group_by(census_tract) |>
  summarize(
    applications = n(),
    population = mean(tract_population),
    minority = mean(tract_minority_population_percent),
    income = mean(tract_to_msa_income_percentage),
    median_value = median(property_value_1000, na.rm = TRUE)
  )
```

## Modeling Property Values vs. Minority

```{r}
#| label: scatterplot of median property values vs minority

ggplot(data = fullerton_tracts,
       mapping = aes(
         x = minority,
         y = median_value
       )) +
  geom_point() +
  geom_smooth(method = "lm", color = "darkblue", 
              linetype = "dashed", se = FALSE) +
  labs(
    title = "Housing Values in Fullerton Census Tracts",
    x = "Percentage of Minorities in Tract",
    y = "Median Property Value (1000s of $)"
  )
```

```{r}
#| label: build model value vs minority

lm_medvalue_minority <- lm(
  median_value ~ minority,
  data = fullerton_tracts
)

lm_medvalue_minority |>
  tidy()
```

### Your Turn

1.  Use the output to write out the equation of the least-squares regression line.

2. Write a sentence to interpret the slope of the regression line. It may help to remember that the observations are census tracts, the units of `minority` are percentage points (% minority) and the units of `median_value` are thousands of dollars.

3. Does this interpretation make sense in the context of the collected data? Why or why not?

4. Write a sentence to interpret the intercept of the regression line.

5. Does this interpretation make sense in the context of the collected data? Why or why not?

## Modeling Property Values vs. Income

### Your Turn

1. Reproduce the scatterplot from the Summarizing Numerical Data Activity with the median property value on the y-axis and the median income (as a percentage of median income in the LA area). Add the least-squares regression line to the scatterplot.

2. Using the `lm` function, fit a linear regression model describing the relationship between the two variables.

3. Using the `tidy` function, obtain the coefficient table. Use the output to write out the equation of the least-squares regression line.

