---
title: "Fitting and Summarizing Linear Models: Rent Prices in Southern California"
author: "Your Name Here!"
format: html
editor: visual
---

In this activity, our goal in this activity is to build a model to predict the listed rent price of a house (in 2022) based on its square footage.

The SoCalRent1 dataset contains information about 183 houses or condos available to rent within a "reasonable distance" of Fullerton in May 2022.

The SoCalRent2 dataset contains information about an additional 117 houses or condos available to rent during the same time period.

See the SocalRent data dictionary for descriptions of the variables.

## Load Packages, Import Data, Massage Data

```{r}
#| label: do this first
here::i_am("Activities/Summarizing-Linear-Models-SoCalRent.qmd")

library(tidyverse)
library(broom)

SoCalRent1 <- read_csv(here::here("Data/SoCalRent1.csv"))
SoCalRent2 <- read_csv(here::here("Data/SoCalRent2.csv"))
```

## Your Turn: Data Wrangling and Visualization

We will *only* do our data wrangling and visualization with the `SoCalRent1` dataset. We will "hold out" the `SocalRent2` dataset to assess how well our linear regression model predicts *new* data.

1.  Create a plot showing the relationship between rent price and square footage.

2.  Write 1-2 sentences describing what you notice about the relationship.

3.  Which points appear to be outliers, and why (unusually small/large, unusually (in)expensive, and/or unusually (in)expensive compared to other houses of similar size)? Investigate the outliers, make a judgment call about which (if any) outliers should be thrown out before fitting the model, and justify that judgment call based on your investigation.

4.  Convert your judgment call(s) into logical expression(s) based on variables in the dataset, then use those logical expressions to `filter` the dataset to include only the houses you actually want to model. Call the resulting dataset `SoCalRent_train`.

```{r}
#| label: create SoCalRent_train

# We will discuss as a class to make sure we all filter the same way before creating this dataset
```

## Building the Model

```{r}
#| label: build model Price vs SqFt

# Code with me to build the model and obtain the coefficient estimates

```

1.  Use the output to write out the equation of the least-squares regression line.

2.  Write a sentence to interpret the slope of the regression line.

3.  Does this interpretation make sense in the context of the collected data? Why or why not?

4.  Write a sentence to interpret the intercept of the regression line.

5.  Does this interpretation make sense in the context of the collected data? Why or why not?

## Making Predictions

1.  Perform the same `filter`ing on the `SoCalRent2` dataset that you did on the `SoCalRent1` dataset. Call this new dataset `SoCalRent_test`.

```{r}
#| label: filter SoCalRent2


```

2.  Predict the rent price of every house in the `SoCalRent_test` dataset.

```{r}
#| label: make predictions

# Code with me to get the predictions
```

3.  Produce a histogram of the prediction errors (residuals) for the `SoCalRent_test` dataset.

```{r}
#| label: histogram of residuals

# Code with me to produce the histogram
```

Briefly describe the distribution of the residuals. Does this histogram suggest anything concerning about the accuracy of the predictions? Why or why not?

4.  Produce a scatterplot with the predicted (fitted) prices on the x-axis and the residuals on the y-axis. Add a linear and/or nonlinear trend to the plot (using `geom_smooth`). Does this scatterplot suggest anything concerning about the accuracy of the predictions? Why or why not?

## Regression with a Categorical Predictor

1.  Write 1-2 sentences describing what you notice about the relationship between house type and rent price in the `SoCalRent_train` dataset. You should support your description with at least one graph and one grouped numerical summary.

2.  Fit a model predicting price from house type on the `SoCalRent_train` data.

3.  Obtain the coefficient table using the `tidy()` function, and write out the equation of the least-squares regression line.

4.  Write a sentence to interpret the slope of the regression line.

5.  Does this interpretation make sense in the context of the collected data? Why or why not?

6.  Write a sentence to interpret the intercept of the regression line.

7.  Does this interpretation make sense in the context of the collected data? Why or why not?

8.  Using the `augment` function, predict the rent price of every house in the `SoCalRent_test` dataset. What do you notice about these predictions? Does (should) it surprise you?
