---
title: "Summarizing Numerical Data with R"
format: revealjs
editor: visual
execute:
  echo: TRUE
---

## Steps 1 and 2: Load Packages and Import Data

```{r}
#| label: load packages 
library(tidyverse) 
```

```{r}
#| label: import data with here
alz <- readr::read_csv(here::here("Data", "alzheimers.csv"))
```

## Distribution of a Variable

-   The distribution of a variable indicates:

    -   What values were observed

    -   How often each value was observed

-   For categorical data, this is represented by:

    -   The levels of the variable

    -   The count or proportion of observations at each level

## Distribution of a Numerical Variable

-   The distribution of a numerical variable is more complicated to describe

-   We consider "3+1" properties of the distribution

    -   **Center**: where the values tend to be on the number line

    -   **Variability**: how spread out the values tend to be along the number line

    -   **Shape**: where the values tend to be *relative to each other*
    
    -   **Outliers**: the possible presence of unusual values

## Describing Variables in Context

Let's look at the distribution of `gmv`, representing the amount of gray matter (volume of cell bodies) in a person's brain

-   **Center**: What would be an "average" or "typical" `gmv` value?

-   **Variability**: How much variation do we see in `gmv` across patients in the sample?

-   **Shape**: Relative to the rest of the sample, do people tend to have higher/lower `gmv` values or to be somewhere in the middle? Are there multiple very common values?
    
-   **Outliers**: Does any patient have an unusually high or low `gmv` value?


## Center: Mean and Median

- The mean is the *average value* of the variable

- The median is the value of the *average observation* (middle value)

```{r}
#| label: mean and median

alz |>
  summarize(
    mean_gmv = mean(gmv),
    median_gmv = median(gmv)
  )
```
## Center and Variability Together

-   When we choose *mean* as our measure of center, the typical associated measure of variability is *standard deviation*

    -   Think of standard deviation as the "average distance from the mean"

-   When we choose *median* as our measure of center, the typical associated measure of variability is *interquartile range* (IQR)

    -   Distance between minimum and maximum values in the *middle 50%* of the dataset

## Center and Variability with Code

```{r}
#| label: mean and sd

alz |>
  summarize(
    mean_gmv = mean(gmv),
    sd_gmv = sd(gmv)
  )
```

```{r}
#| label: median and IQR

alz |>
  summarize(
    median_gmv = median(gmv),
    iqr_gmv = IQR(gmv)
  )
```

## Center and Variability in Context

```{r}
#| label: all the summaries-1
#| echo: FALSE
#| eval: TRUE
alz |>
  summarize(
    mean_gmv = mean(gmv),
    sd_gmv = sd(gmv)
  )
```

Patients had an average gray matter volume of about 573 cc. A typical patient had a gray matter volume 61 cc away from this average.

```{r}
#| label: all the summaries-2
#| echo: FALSE
#| eval: TRUE

alz |>
  summarize(
    median_gmv = median(gmv),
    iqr_gmv = IQR(gmv)
  )
```

The average patient had a gray matter volume of about 569 cc. The middle 50% of patient gray matter volumes spanned a range of 98 cc.

## Five Number Summary

- The *five number summary* divides the dataset into four quarters

    -   Each quarter contains 25% of the data

```{r}
#| label: five number summary with summarize

alz |>
  summarize(
    min_gmv = min(gmv), # minimum
    q1_gmv = quantile(gmv, 0.25), # lower quartile or Q1
    med_gmv = median(gmv), # median
    q3_gmv = quantile(gmv, 0.75), # upper quartile or Q3
    max_gmv = max(gmv) # maximum
  )
```

## Five Number Summary in Context

```{r}
#| label: five number summary-2
#| echo: FALSE
#| eval: TRUE
alz |>
  summarize(
    min_gmv = min(gmv), # minimum
    q1_gmv = quantile(gmv, 0.25), # lower quartile
    med_gmv = median(gmv), # median
    q3_gmv = quantile(gmv, 0.75), # upper quartile
    max_gmv = max(gmv) # maximum
  )
```

-   All patients had gray matter volume between 441 cc (min) and 721 cc (max)

-   The middle 50% of patients had gray matter volume between 523 (Q1) and 620 cc (Q3)

-   The average patient had a gray matter volume of 569 cc

## Describing Shape and Outliers

- Typically done visually

- Two main options for graphing the distribution:

    -   Histogram
    
    -   Box plot
    
## Creating a Histogram

::: columns
::: {.column width="55%"}
```{r}
#| label: hist1-1
#| echo: TRUE
#| eval: FALSE
# Step 1: Set up the graph
ggplot(data = alz, 
       mapping = aes(x = gmv)) +
# Step 2: Add the plot
  geom_histogram() +
# Step 3: Add features to the plot
  labs(x = "Gray Matter Volume (cc)",
       y = "Number of Patients") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::

::: {.column width="45%"}
```{r}
#| label: hist1-2
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
ggplot(data = alz, 
       mapping = aes(x = gmv)) +
# Step 2: Add the plot
  geom_histogram() +
# Step 3: Add features to the plot
  labs(x = "Gray Matter Volume (cc)",
       y = "Number of Patients") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::
:::

## Reading a Histogram

-   A histogram is a bar graph whose x-axis is the number line

-   We create a categorical variable by "binning" the values into intervals along the number line

    -   By convention, intervals are equally spaced
    
    -   However, we have control over their center and width
    
-   We then place the bar over that interval on the x-axis

    -   This means the bars are contiguous (touching)
    
    -   A gap between the bars really means a bar of height 0

## Customizing a Histogram

::: columns
::: {.column width="55%"}
```{r}
#| label: hist2-1
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "5-8"
# Step 1: Set up the graph
ggplot(data = alz, 
       mapping = aes(x = gmv)) +
# Step 2: Add the plot
  geom_histogram(fill = "darkviolet", 
                 color = "black",
                 center = 500,
                 binwidth = 20) +
# Step 3: Add features to the plot
  labs(x = "Gray Matter Volume (cc)",
       y = "Number of Patients") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::

::: {.column width="45%"}
```{r}
#| label: hist2-2
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
# Step 1: Set up the graph
ggplot(data = alz, 
       mapping = aes(x = gmv)) +
# Step 2: Add the plot
  geom_histogram(fill = "darkviolet", 
                 color = "black",
                 center = 500,
                 binwidth = 20) +
# Step 3: Add features to the plot
  labs(x = "Gray Matter Volume (cc)",
       y = "Number of Patients") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::
:::

## Describing Shape with Histograms: Skewed Right

::: columns
::: {.column width="55%"}
```{r}
#| label: skewed right
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
# Step 1: Set up the graph
ggplot(data = alz, 
       mapping = aes(x = csf)) +
# Step 2: Add the plot
  geom_histogram(fill = "darkviolet", 
                 color = "black",
                 center = 275,
                 binwidth = 50) +
# Step 3: Add features to the plot
  labs(x = "Cerebrospinal Fluid Volume (cc)",
       y = "Number of Patients") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::

::: {.column width="45%"}

-   Major peak on the left side of the graph
    
-   Long "tail" to the right side
    
-   Looks similar to tracing your right foot

:::
:::

## Describing Shape with Histograms: Skewed Left

::: columns
::: {.column width="55%"}
```{r}
#| label: skewed left
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
# Step 1: Set up the graph
ggplot(data = alz, 
       mapping = aes(x = mmse)) +
# Step 2: Add the plot
  geom_histogram(fill = "darkviolet", 
                 color = "black",
                 center = 29,
                 binwidth = 2) +
# Step 3: Add features to the plot
  labs(x = "Mini-Mental State Examination Score",
       y = "Number of Patients") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::

::: {.column width="45%"}

-   Major peak on the right side of the graph
    
-   Long "tail" to the left side
    
-   Looks similar to tracing your left foot

:::
:::

## Describing Shape with Histograms: Roughly Symmetric

::: columns
::: {.column width="55%"}
```{r}
#| label: roughly symmetric
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
# Step 1: Set up the graph
ggplot(data = alz, 
       mapping = aes(x = wmv)) +
# Step 2: Add the plot
  geom_histogram(fill = "darkviolet", 
                 color = "black",
                 center = 480,
                 binwidth = 40) +
# Step 3: Add features to the plot
  labs(x = "White Matter Volume (cc)",
       y = "Number of Patients") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::

::: {.column width="45%"}

-   Left and right halves of histogram look roughly like mirror images
    
-   With real data, we shouldn't expect perfect symmetry
    
-   If it's difficult to tell which direction the skew is in, call it roughly symmetric

:::
:::

## Identifying Outliers on Histograms

::: columns
::: {.column width="55%"}
```{r}
#| label: weight outlier
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
# Step 1: Set up the graph
ggplot(data = alz, 
       mapping = aes(x = weight)) +
# Step 2: Add the plot
  geom_histogram(fill = "darkviolet", 
                 color = "black",
                 center = 75,
                 binwidth = 10) +
# Step 3: Add features to the plot
  labs(x = "Weight (kg)",
       y = "Number of Patients") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::

::: {.column width="45%"}

-   Unusually small or large values
    
    -   Judgment call how "unusual" 
    
-   Look for values on the left or right side separated from the bulk of the data by a gap of several "height 0" bars 
    
:::
:::

## Creating a Boxplot

::: columns
::: {.column width="55%"}
```{r}
#| label: boxplot1-1
#| echo: TRUE
#| eval: FALSE
# Step 1: Set up the graph
ggplot(data = alz, 
       mapping = aes(x = gmv)) +
# Step 2: Add the plot
  geom_boxplot(fill = "blue") +
# Step 3: Add features to the plot
  labs(x = "Gray Matter Volume (cc)",
       y = "") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::

::: {.column width="45%"}
```{r}
#| label: boxplot1-2
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
ggplot(data = alz, 
       mapping = aes(x = gmv)) +
# Step 2: Add the plot
  geom_boxplot(fill = "blue") +
# Step 3: Add features to the plot
  labs(x = "Gray Matter Volume (cc)",
       y = "") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::
:::

## Reading a Boxplot

-   "Box" spans from lower to upper quartile

    -   Median as a thick line in the middle of the box

-   "Whiskers" extend to smallest and largest points that are *not outliers*

-   Outliers are marked as individual dots

## Convention: Outliers on Boxplots

::: columns
::: {.column width="55%"}
```{r}
#| label: boxplot with outliers
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
# Step 1: Set up the graph
ggplot(data = alz, 
       mapping = aes(x = weight)) +
# Step 2: Add the plot
  geom_boxplot(fill = "blue") +
# Step 3: Add features to the plot
  labs(x = "Weight (kg)",
       y = "") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::

::: {.column width="45%"}

- We compute invisible fences:

$$
Lower \ Fence = Q_1 - 1.5 \times IQR
$$

$$
Upper \ Fence = Q_3 + 1.5 \times IQR
$$

- Any data value *not* between the fences is considered an outlier

:::
:::

## Comparing Groups

Let's compare the distribution of `gmv` between Alzheimer's patients and healthy controls

- **Center**: What number do the `gmv` values tend to be around? Does this change depending on the diagnosis?
  
- **Variability**: How spread out are the `gmv` values? Does this change depending on the diagnosis?

- **Shape**: Does the distribution of `gmv` "look the same" in each group, only potentially shifted/scaled?

- **Outliers**: For each diagnosis group, are there any values that are unusual for that group? If so, what makes them unusual?

Always note anything else that caught your attention, even if it doesn't neatly map to one of these properties

## Comparing Groups with Numerical Summaries

```{r}
#| label: five number summary by group
#| code-line-numbers: "2-3"
alz |>
  group_by(diagnosis) |>
  summarize(
    mean_gmv = mean(gmv),
    median_gmv = median(gmv),
    sd_gmv = sd(gmv),
    IQR_gmv = IQR(gmv)
  )
```

## Comparing Groups with Boxplots

::: columns
::: {.column width="55%"}
```{r}
#| label: multiple boxplots-1
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "2-4"
ggplot(data = alz, 
       mapping = aes(x = diagnosis, 
                     y = gmv)) +
  geom_boxplot() +
  labs(x = "Diagnosis",
       y = "Gray Matter Volume (cc)") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```

:::

::: {.column width="45%"}
```{r}
#| label: multiple boxplots-2
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
ggplot(data = alz, 
       mapping = aes(x = diagnosis, 
                     y = gmv)) +
  geom_boxplot() +
  labs(x = "Diagnosis",
       y = "Gray Matter Volume (cc)") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::
:::

- Compare center and variability between groups by reading median and IQR off the boxplots

## Customizing Boxplots by Group

::: columns
::: {.column width="55%"}
```{r}
#| label: multiple boxplots-2-1
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "8-11"
ggplot(data = alz, 
       mapping = aes(x = diagnosis, 
                     fill = diagnosis, 
                     y = gmv)) +
  geom_boxplot() +
  labs(x = "Diagnosis",
       y = "Gray Matter Volume (cc)") +
  scale_x_discrete(labels = c("Alzheimer's", "Normal")) +
  scale_fill_manual(name = "Diagnosis",
                    labels = c("Alzheimer's", "Normal"), 
                    values = c("red4","red")) +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```

:::

::: {.column width="45%"}
```{r}
#| label: multiple boxplots-2-2
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
ggplot(data = alz, 
       mapping = aes(x = diagnosis, 
                     fill = diagnosis, 
                     y = gmv)) +
  geom_boxplot() +
  labs(x = "Diagnosis",
       y = "Gray Matter Volume (cc)") +
  scale_x_discrete(labels = c("Alzheimer's", "Normal")) +
  scale_fill_manual(name = "Diagnosis",
                    labels = c("Alzheimer's", "Normal"), 
                    values = c("red4","red")) +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::
:::

## Comparing Shape Between Groups

::: columns
::: {.column width="55%"}
```{r}
#| label: multi-hist-1
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "9"
ggplot(data = alz, 
       mapping = aes(x = gmv)) +
  geom_histogram(fill = "darkviolet", 
                 color = "black",
                 center = 500,
                 binwidth = 20) +
  labs(x = "Gray Matter Volume (cc)",
       y = "Number of Patients") +
  facet_wrap(vars(diagnosis)) +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::

::: {.column width="45%"}
```{r}
#| label: multi-hist-2
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
ggplot(data = alz, 
       mapping = aes(x = gmv)) +
  geom_histogram(fill = "darkviolet", 
                 color = "black",
                 center = 500,
                 binwidth = 20) +
  labs(x = "Gray Matter Volume (cc)",
       y = "Number of Patients") +
  facet_wrap(vars(diagnosis)) +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::
:::

## Relationships Between Numerical Variables

::: columns
::: {.column width="55%"}
```{r}
#| label: scatterplot-1
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "2-4"
ggplot(data = alz, 
       mapping = aes(x = age, 
                     y = gmv)) +
  geom_point(color = "navy", size = 2) +
  labs(x = "Age (years)",
       y = "Gray Matter Volume (cc)") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```

:::

::: {.column width="45%"}
```{r}
#| label: scatterplots-2
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
ggplot(data = alz, 
       mapping = aes(x = age, 
                     y = gmv)) +
  geom_point(color = "navy", size = 2) +
  labs(x = "Age (years)",
       y = "Gray Matter Volume (cc)") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::
:::

## Describing Relationships Between Numerical Variables

- **Direction**: What tends to happen to `gmv` as people get older?

- **Form**: What functions (if any) might model the trend well? 

- **Strength**: How tightly does the relationship follow the trend you identified?

- **Outliers**: Are there any unusual points? If so, what makes them unusual?

- Is there anything else that caught your attention?

## Visualizing Linear Trends

::: columns
::: {.column width="55%"}
```{r}
#| label: lm-1
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "5-6"
ggplot(data = alz, 
       mapping = aes(x = age, 
                     y = gmv)) +
  geom_point(color = "navy", size = 2) +
  geom_smooth(method = "lm", color = "black", 
              se = FALSE) +
  labs(x = "Age (years)",
       y = "Gray Matter Volume (cc)") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```

:::

::: {.column width="45%"}
```{r}
#| label: lm-2
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
ggplot(data = alz, 
       mapping = aes(x = age, 
                     y = gmv)) +
  geom_point(color = "navy", size = 2) +
  geom_smooth(method = "lm", color = "black", 
              se = FALSE) +
  labs(x = "Age (years)",
       y = "Gray Matter Volume (cc)") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::
:::

## Visualizing Nonlinear Trends

::: columns
::: {.column width="55%"}
```{r}
#| label: nlm-1
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "5-6"
ggplot(data = alz, 
       mapping = aes(x = age, 
                     y = gmv)) +
  geom_point(color = "navy", size = 2) +
  geom_smooth(method = "loess", color = "black", 
              se = FALSE) +
  labs(x = "Age (years)",
       y = "Gray Matter Volume (cc)") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```

:::

::: {.column width="45%"}
```{r}
#| label: nlm-2
#| echo: FALSE
#| eval: TRUE
#| fig-height: 8
ggplot(data = alz, 
       mapping = aes(x = age, 
                     y = gmv)) +
  geom_point(color = "navy", size = 2) +
  geom_smooth(method = "loess", color = "black", 
              se = FALSE) +
  labs(x = "Age (years)",
       y = "Gray Matter Volume (cc)") +
  theme(
    axis.title = element_text(size = 24),
    axis.text = element_text(size = 16)
  )
```
:::
:::
